{\small} High Performance P$^{3}$M N-body code: CUBEP3M

### 2012-06-06 Structure.tex
Modifications after meeting with Ilian and Garrelt in Stockholm. Lots of edits in the text, new figures, included halo mass function for extended pp_range (nice agreement with ST), added Hugh Merz comments about BlueGene.

As mentioned in the Introduction section, {\small CUBEP3M} is a {\small FORTRAN90} 
N-body code that solves Poisson's equation on a two-level mesh, 
with sub-cell accuracy thanks to particle-particle interactions. 
The code has extensions that departs from this basic scheme, and
we shall come back to these later, but for the moment, we adopt the 
standard configuration. 
The long range component of the gravity force is solved on the coarse grid, 
and is global in the sense that the calculations require knowledge about the full simulated volume.
The short range force and the particle-particle interactions are computed in parallel on local volumes\footnote{To make this possible, the fine grid arrays are constructed such as to support parallel reading and writing. In practice, this is done by adding an additional dimension to the relevant arrays, such that each {\small CPU} accesses a unique memory location.} or {\it tiles}. The force matching of the two meshes is performed by introducing a cutoff in the two force kernel. The value of $r_{c}=16$ fine cells was chosen such as to balance the communication 
overhead between processes and the accuracy of the match between the two meshes.

The computation of the short range force requires each tile to store the fine grid density in a buffer surface around the physical volume it is assigned. The thickness of this surface must be larger than the cutoff length (of 16 cells), and we find that a 24 cells deep buffer
is a good compromise between memory usage and accuracy.
When it comes to finding haloes at run time can this buffer create a problem, because a large halo located close to the boundary
can have a radius larger than the buffer zone, in which case it would be truncated, thus falling in the wrong mass bin, producing an incorrect center of mass, etc. 
It could then be desirable to increase the buffer zone around each tile, at the cost of a loss of memory dedicated to the actual physical volume.

@@

Because the coarse grid arrays require $4^3$ times less memory per node, 
the bulk of the foot-print is concentrated in a handful of fine grid arrays.
In addition, some of these are required for intermediate steps of the calculations only, 
hence it is possible to hide some of the coarse grid arrays\footnote{ This memory recycling is done with `equivalence' statements in {\small F90}}.   
We present here the largest arrays used by the code:
\begin{enumerate}
\item{{\tt xv} stores the position and velocity of each particle} 
\item{{\tt ll} stores the linked-list that accelerate the access to particles in each coarse grid cell}
\item{{\tt rho\_f} and {\tt cmplx\_rho\_f} store 
the local fine grid density  in real and Fourier space respectively}
\item{{\tt force\_f} stores the force of gravity (short range only) along the three Cartesian directions}
\item{{\tt kern\_f} stores the fine grid force kernel in the three directions}
\item{{\tt PID} stores the unique particle identification tags.}
\end{enumerate}
The particle ID is a feature that can be switched off by removing a compilation flag, 
and allows to optimize the code for higher resolution configurations.
%\subsection{Code overview}
%\label{subsec:overview}

@@

Each iteration of the main loop starts with the {\tt timestep} subroutine, 
which proceeds to a determination of the redshift jump by comparing the step size constraints from each
force components and from the scale factor.
The cosmic expansion is found by Taylor expanding Friedmann's equation up to the third order in the scale factor,
and can accommodate constant or running equation of state of dark energy.
The force of gravity is then solved  in the {\tt particle\_mesh} subroutine,
which first updates the positions and velocities of the dark matter particles, exchange with neighbouring nodes those that have exited to volume,
creates a new linked list, then solve Poisson's equation.  This subroutine is conceptually identical to that of {\small PMFAST}, 
with the exception  that {\small CUBEP3M} decomposes the volume into cubes (as opposed to slabs). 
The loop over tiles and the particle exchange are thus performed in three dimensions.
{\bf (Should I mention leapfrog here or anywhere else?)}
When the short range and pp forces have been calculated on all tiles, the code exits the parallel {\small OPENMP} loop
and proceeds to the long range. This section of the code is also parallelized in many occasions, but, unfortunately, the current {\small MPI-FFTW}
do not allow multi-threading. There is thus an inevitable loss of efficiency during each global Fourier transforms, during which
only the single core {\small MPI} process is active\footnote{Other libraries such as {\small P3DFFT} ({\tt http://code.google.com/p/p3dfft/}) currently permit
 an extra level of parallelization, and it is our plan to migrate to one of these in the near future.}.

\begin{figure}
\begin{verbatim}
program cubep3m
   call initialize
   do
       call timestep
       call particle_mesh
       if(checkpoint_step) then
          call checkpoint
       elseif(last_step)
          exit
       endif
   enddo
   call finalize
end program cubep3m
\end{verbatim}
\caption{Overall structure of the code.}
\label{fig:structure}
\end{figure}

\begin{figure}
\begin{verbatim}
subroutine particle_mesh
   call update_position
   call link_list
   call particle_pass
   !$omp parallel do
   do tile = 1, tiles_node
      call rho_f_ngp
      call cmplx_rho_f
      call kernel_multiply_f
      call force_f
      call update_velocity_f
      if(pp = .true.) then       
         call link_list_pp
         call force_pp
         call update_velocity_pp
         if(extended_pp = .true.) then
            call link_list_pp_extended
            call force_pp_extended
            call update_velocity_pp_extended       
         endif
      endif
   end do
   !$omp end parallel do
   call rho_c_ngp
   call cmplx_rho_c
   call kernel_multiply_c
   call force_c
   call update_velocity_c      
   delete_buffers
end subroutine particle_mesh
\end{verbatim}
\caption{Overall structure of the two-level mesh algorithm. We have included the section that concerns the extended pp force calculation to illustrate that it follows the same basic logic. We mention here that the three linked list arrays are distinct entities, and their structure differ slightly.  }
\label{fig:particle_mesh}
\end{figure}

@@

Other considerations that need to be considered is that any decomposition geometry somehow 
limits permissible grid sizes, and that volume evenly decomposed into cubic sub-sections 
--  such as {\small CUBEP3M} -- requires the number of MPI processes to be a perfect cube.
Also,  since the decomposition is volumetric, as opposed to density dependent -- 
it suffers from load imbalance when probing deep into the non-linear regime.

>>>>

As mentioned in the Introduction section, {\small CUBEP3M} is a {\small FORTRAN90} 
N-body code that solves Poisson's equation on a two-level mesh, 
with sub-cell accuracy thanks to particle-particle interactions. 
The code has extensions that departs from this basic scheme, and
we shall come back to these later, but for the moment, we adopt the 
standard configuration. 
The long range component of the gravity force is solved on the coarse grid, 
and is global in the sense that the calculations require knowledge about the full simulated volume.
The short range force and the particle-particle interactions are computed in parallel on
a second level of cubical decomposition of the local volumes, the {\it tiles}. To make this possible, the fine grid arrays are constructed such as to support parallel memory access. In practice, this is done by adding an additional dimension to the relevant arrays, such that each {\small CPU} accesses a unique memory location. The force matching between the two meshes is performed by introducing a cutoff length, $r_{c}$, in the definition of the two force kernels. The value of $r_{c}=16$ fine cells was found to balance the communication 
overhead between processes and the accuracy of the match between the two meshes.

The computation of the short range force requires each tile to store the fine grid density of a region that includes a buffer surface around the physical volume it is assigned. The thickness of this surface must be larger than $r_{c}$, and we find that a 24 cells deep buffer
is a good compromise between memory usage and accuracy.
This is basically to fully compensate for the coarse mesh calculations, whose CIC interpolation scheme 
reaches two coarse cells deep beyond the cutoff.

When it comes to finding haloes at run time, this buffer can create a problem, because a large object located close to the boundary can have a radius larger than the buffer zone, in which case it would be truncated and be assigned a wrong mass, center of mass, etc. 
It could then be desirable to increase the buffer zone around each tile, at the cost of a loss of memory dedicated to the actual physical volume, and the code is designed to allow for such a flexibility.

@@

Because the coarse grid arrays require $4^3$ times less memory per node, 
it does not contribute much to the total memory requirement, and the bulk of the foot-print is 
concentrated in a handful of fine grid arrays.
Some of these are only required for intermediate steps of the calculations, 
hence it is possible to hide therein many coarse grid arrays\footnote{ This memory recycling is done with `equivalence' statements in {\small F90}}.   
We present here the largest arrays used by the code:
\begin{enumerate}
\item{{\tt xv} stores the position and velocity of each particle} 
\item{{\tt ll} stores the linked-list that accelerate the access to particles in each coarse grid cell}
\item{{\tt rho\_f} and {\tt cmplx\_rho\_f} store 
the local fine grid density  in real and Fourier space respectively}
\item{{\tt force\_f} stores the force of gravity (short range only) along the three Cartesian directions}
\item{{\tt kern\_f} stores the fine grid force kernel in the three directions}
\item{{\tt PID} stores the unique particle identification tags.}
\end{enumerate}
The particle ID is a feature that can be switched off by removing a compilation flag, 
and allows to optimize the code for higher resolution configurations.
%\subsection{Code overview}
%\label{subsec:overview}

@@

Each iteration of the main loop starts with the {\tt timestep} subroutine, 
which proceeds to a determination of the redshift jump by comparing the step size constraints from each
force components and from the scale factor.
The cosmic expansion is found by Taylor expanding Friedmann's equation up to the third order in the scale factor,
and can accommodate constant or running equation of state of dark energy.
The force of gravity is then solved  in the {\tt particle\_mesh} subroutine,
which first updates the positions and velocities of the dark matter particles, exchange with neighbouring nodes those that have exited to volume,
creates a new linked list, then solve Poisson's equation.  This subroutine is conceptually identical to that of {\small PMFAST}, 
with the exception  that {\small CUBEP3M} decomposes the volume into cubes (as opposed to slabs). 
The loop over tiles and the particle exchange are thus performed in three dimensions.
When the short range and pp forces have been calculated on all tiles, the code exits the parallel {\small OPENMP} loop
and proceeds to the long range. This section of the code is also parallelized on many occasions, but, unfortunately, the current {\small MPI-FFTW}
do not allow multi-threading. There is thus an inevitable loss of efficiency during each global Fourier transforms, during which
only the single core {\small MPI} process is active\footnote{Other libraries such as {\small P3DFFT} ({\tt http://code.google.com/p/p3dfft/}) currently permit
 an extra level of parallelization, and it is our plan to migrate to one of these in the near future.}.
 As in {\small PMFAST}, the particle position and velocity updates are performed in a leap frog scheme \citep{1981csup.book.....H}.

\begin{figure}
\begin{verbatim}
program cubep3m
   call initialize
   do
       call timestep
       call particle_mesh
       if(checkpoint_step) then
          call checkpoint
       elseif(last_step)
          exit
       endif
   enddo
   call finalize
end program cubep3m
\end{verbatim}
\caption{Overall structure of the code simplified for readability.}
\label{fig:structure}
\end{figure}

\begin{figure}
\begin{verbatim}
subroutine particle_mesh
   call update_position + apply random offset
   call link_list
   call particle_pass
   !$omp parallel do
   do tile = 1, tiles_node
      call rho_f_ngp
      call cmplx_rho_f
      call kernel_multiply_f
      call force_f
      call update_velocity_f
      if(pp = .true.) then       
         call link_list_pp
         call force_pp
         call update_velocity_pp
         if(extended_pp = .true.) then
            call link_list_pp_extended
            call force_pp_extended
            call update_velocity_pp_extended       
         endif
      endif
   end do
   !$omp end parallel do
   call rho_c_ngp
   call cmplx_rho_c
   call kernel_multiply_c
   call force_c
   call update_velocity_c      
   delete_buffers
end subroutine particle_mesh
\end{verbatim}
\caption{Overall structure of the two-level mesh algorithm. We have included the section that concerns the standard pp and the extended pp force calculation, to illustrate that they follows similar linked-list logic. }
\label{fig:particle_mesh}
\end{figure}

@@

Other considerations that need to be considered is that any decomposition geometry somehow 
limits permissible grid sizes, and that volume evenly decomposed into cubic sub-sections 
--  such as {\small CUBEP3M} -- requires the number of {\small MPI} processes to be a perfect cube.
Also,  since the decomposition is volumetric, as opposed to density dependent -- 
it suffers from load imbalance when probing deep into the non-linear regime.

### 2012-06-06 Systematics.tex
Modifications after meeting with Ilian and Garrelt in Stockholm. Lots of edits in the text, new figures, included halo mass function for extended pp_range (nice agreement with ST), added Hugh Merz comments about BlueGene.

One of the most reliable way to assess the simulation's accuracy at evolving particles
is to measure the density power spectrum at late time, and compare to non-linear prediction. 
For an over-density field $\delta({\bf \it x})$, the power spectrum is extracted from the two point function in Fourier space as:
\begin{eqnarray}
\langle | \delta ({\bf \it k}) \delta ({\bf \it k'}) | \rangle = (2\pi)^{3}P(k)\delta_{D}({\bf \it k'} - {\bf \it k})
\label{eq:power}
\end{eqnarray}
where the angle bracket corresponds to an ensemble (or volume) average.
We plot in Fig. \ref{fig:power_highres} the dimensionless power spectrum in a RANGER4000 simulation and
observe that the agreement with the non-linear prediction of \cite{Lewis:1999bs} is at the per cent level over a large dynamical range.
The drop of power at high-$k$ is caused by the finite resolution, and the fluctuations at low-$k$ are caused by the white noise imposed
in the initial conditions.

\begin{figure*}%[ht]
  \begin{center}
    \includegraphics[width=5.2in]{graphs/power_highres.eps}
  \caption{Dark matter power spectrum, measured at $z=0$ in a RANGER4000 simulation.
    \label{fig:power_highres}}
\end{center}
\end{figure*}

The actual force of gravity in the p$^3$m algorithm,
as felt by a single particle, is presented in Fig. \ref{fig:den_force_ppext0}.
This shows the force versus distance -- in fine cell units -- a calculation that was performed in a CITA256 realization in two steps: 
1- we compute the force on each particle in a given distribution, as done at every time step.
2- we remove a selected particle, compute the force again on all particles, and record on a file the 
force difference (before  and after the removal) as a function of the distance to the `hole'.

Particles in the same fine cell follow the exact $1/r^{2}$ curve. The scatter at 
  distances of the order of the fine grid is caused by the NGP interpolation scheme:
  Particles in adjacent fine cells can be actually very close, as seen in the upper left region of this plot,
  but feel the mesh force at grid cell distances,
  creating up to an order of magnitude loss in the force.
As the separation approaches a tenth of the box size or so, the force on the coarse mesh deviates
from Newton's law in order to preserve periodic boundary conditions. 
This feature, however, can be turned off at compilation time.

Fig. \ref{fig:den_force_fracErr} shows the fractional error on the force along the radial and tangential directions.
It is larger than in {\small PMFAST} at grid scales, largely due to the fact that the fine mesh force is performed with an NGP interpolation scheme -- as opposed to CIC -- which shows a larger scatter about the theoretical value. 
As discuss in the next section, to minimize this effect, we apply a random offset to all the particles,
such that this undesired scatter is average out over a few time steps.

\begin{figure}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/densityForce_ppext=0.eps}
  \caption{Gravity force in of the p$^3$m algorithm, versus distance in fine mesh cell units, compared with the exact $1/r^{2}$ law.
    This particular calculation was obtained in a CITA256 realization. 
    \label{fig:den_force_ppext0}}
\end{center}
\end{figure}

\begin{figure}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/densityForce_fracErr.eps}
  \caption{Fractional error on the force in of the p$^3$m algorithm, in the radial (top) and tangential(bottom) directions.
    \label{fig:den_force_fracErr}}
\end{center}
\end{figure}

@@

The biggest problem with the straightforward pp force calculation is that the results 
are anisotropic and depend on the location of the fine mesh with respect 
to the particles. As an example, consider two particles on either side of a grid 
cell boundary, experiencing their mutual gravity attraction via the (NGP) mesh force with a 1-grid cell separation.
 If, instead, the mesh was shifted such that they were
within the same cell, they would experience the (larger) pp force. 
This effect is especially pronounced at the early stages of the simulation where
the density is more homogeneous, and leads to mesh artefacts appearing
in the density field. In order to minimize this systematic effect, 
we randomly shift the particle distribution relative to the mesh by a small
amount -- up to 2 fine grid cells in magnitude -- in each
dimension and at each time step.  This adds negligible computational
overhead as it is applied during the particle position update,
and suppresses the mesh behaviour that otherwise grows over multiple time steps.
It is possible to shift back the particles at the end of each time steps,
which prevents a random drift of the whole population, a necessary step 
if one needs to correlate the initial and final positions of the particles for instance,
or for hybrid dark matter -- MHD simulations.

@@

We present in Fig. \ref{fig:disp_mesh} the dramatic impact of removing the random offset in the code.
The power spectrum is completely wrong, caused by the large scatter in the force on the fine mesh.
In this scatter is not averaged over, the error directly adds up at each time step. {\small PMFAST}
did not have this problem since it used CIC interpolation on both meshes.  
This test was performed with CITA128 simulations of very large box size,
which should in principle agree with linear predictions up to the resolution.
The output redshifts are very early, and the upper most curve ($z=10$) was obtained after
about 60 timesteps.
We see that deviations are enormous when the random offset is turned off.

\begin{figure}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/power_w_wo_disp_mesh.eps}
  \caption{Dark matter power spectrum, measured at $z=180$, $100$, $20$ and $10$, in a series of CITA128 realizations that are 1000 Mpc/$h$ per side. The dashed line represent the initial condition power spectrum, the dotted lines are the linear predictions, and  the open circles the standard p$^3$m configuration. 
  The dots were obtained by simply removing the random particle offset that is usually applied at each time step. \label{fig:disp_mesh}}
\end{center}
\end{figure}

@@

It is possible to reduce this effect, and thereby improve significantly 
the accuracy of the code, by modifying the value of $r_{max}$, at the cost of increasing the total number of time steps.
Fig. \ref{fig:ra_max} shows a comparison of late time power spectra of a series of CITA256 realizations that originate from the same initial conditions, 
and used the same random seeds to control the fine mesh shifts (mentioned above): only the value of $r_{max}$ was modified between each run. 
We observe that the impact is mostly located in the non-linear regime, where decreasing the time step to 0.006 
allows the simulation to recover about 30 per cent of dimensionless power at the turn over scale, in this simulation configuration.
This gain is greatly affected by the choice of initial redshift, the resolution, and the box size, and ideally one would make
test runs in order to optimize a given configuration.  
As expected, the {\small CPU} resources required to run these simulations increase rapidly as $r_{max}$ decreases, as seen in Table \ref{table:ra_max}. 
In this test case, reducing further at 0.001 shows only a mild improvement in accuracy, but the increase in time is more than a factor of four.

>>>>

One of the most reliable way to assess the simulation's accuracy at evolving particles
is to measure the density power spectrum at late time, and compare to non-linear prediction. 
For an over-density field $\delta({\bf \it x})$, the power spectrum is extracted from the two point function in Fourier space as:
\begin{eqnarray}
\langle | \delta ({\bf \it k}) \delta ({\bf \it k'}) | \rangle = (2\pi)^{3}P(k)\delta_{D}({\bf \it k'} - {\bf \it k})
\label{eq:power}
\end{eqnarray}
where the angle bracket corresponds to an ensemble (or volume) average.
We plot in Fig. \ref{fig:power_highres} the dimensionless power spectrum in a RANGER4000 simulation and
observe that the agreement with the non-linear prediction of \cite{Lewis:1999bs} is at the per cent level over a large dynamical range.
The drop of power at high-$k$ is caused by the finite resolution, and the fluctuations at low-$k$ are caused by the white noise imposed in the initial conditions. {\b (I should add a vertical line at the k of the coarse mesh for visualization purposes.)}

\begin{figure*}%[ht]
  \begin{center}
    \includegraphics[width=5.2in]{graphs/power_highres.eps}
  \caption{Dark matter power spectrum, measured at $z=0$ in a RANGER4000 simulation,
  compared to the non-linear predictions of {\small CAMB}.
    \label{fig:power_highres}}
\end{center}
\end{figure*}

The actual force of gravity in the P$^3$M algorithm,
as felt by a single particle, is presented in Fig. \ref{fig:den_force_ppext0}.
This shows the force versus distance -- in fine cell units -- a calculation that was performed from a CITA256 realization in two steps: 
1- we compute the force on each particle in a given time step.
2- we remove a selected particle, compute the force again on all particles, and record on file the 
force difference (before  and after the removal) as a function of the distance to the `hole'.

Particles in the same fine cell as the hole follow the exact $1/r^{2}$ curve. The scatter at 
  distances of the order of the fine grid is caused by the NGP interpolation scheme:
  particles in adjacent fine cells can be actually very close, as seen in the upper left region of this plot,
  but still feel the mesh force at grid cell distances,
  creating up to an order of magnitude loss.
As the separation approaches a tenth of the full box size or so, the force on the coarse mesh deviates
from Newton's law in order to preserve periodic boundary conditions. 
This feature, however, can be turned off at compilation time.

Fig. \ref{fig:den_force_fracErr} shows the fractional error on the force along the radial and tangential directions.
It is larger than in {\small PMFAST} at grid scales, largely due to the fact that the fine mesh force is performed with an NGP interpolation scheme -- as opposed to CIC -- which shows a larger scatter about the theoretical value. 
As discuss in the next section, to minimize this effect, we apply a random offset to all the particles,
such that this undesired scatter is averaged out over a few time steps.

\begin{figure}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/densityForce_ppext=0.eps}
  \caption{Gravity force in of the P$^3$M algorithm, versus distance in fine mesh cell units, compared with the exact $1/r^{2}$ law.
    This particular calculation was obtained in a CITA256 realization with  a box size of $500 h^{-1}\mbox{Mpc}$,
    in a single time step. In a full simulation run, the scatter averages over many time steps, thanks to the inclusion of a random offset that is imposed on each particle. 
    \label{fig:den_force_ppext0}}
\end{center}
\end{figure}

\begin{figure}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/densityForce_fracErr.eps}
  \caption{Fractional error on the force in of the P$^3$M algorithm, in the radial (top) and tangential (bottom) directions.
  This scatter is significantly reduced with the random offset, as discussed in section \ref{sec:systematics}.
    \label{fig:den_force_fracErr}}
\end{center}
\end{figure}

@@

The biggest problem with the straightforward pp force calculation is that the results 
are anisotropic and depend on the location of the fine mesh with respect 
to the particles. As an example, consider two particles on either side of a grid 
cell boundary, experiencing their mutual gravity attraction via the fine mesh force with a discretized 1-grid cell separation.
 If, instead, the mesh was shifted such that they were
within the same cell, they would experience the much larger pp force. 
This effect is especially pronounced at the early stages of the simulation where
the density is more homogeneous, and leads to mesh artefacts appearing
in the density field. In order to minimize this systematic effect, 
we randomly shift the particle distribution relative to the mesh by a small
amount -- up to 2 fine grid cells in magnitude -- in each
dimension and at each time step.  This adds negligible computational
overhead as it is applied during the particle position update,
and suppresses the mesh behaviour that otherwise grows over multiple time steps.
It is possible to shift back the particles at the end of each time steps,
which prevents a random drift of the whole population, a necessary step 
if one needs to correlate the initial and final positions of the particles for instance,
or for hybrid dark matter -- MHD simulations.

@@

We present in Fig. \ref{fig:disp_mesh} the dramatic impact of removing the random offset in the code.
The power spectrum is completely wrong, due to the large scatter in the force from the fine mesh.
In this scatter is not averaged over, the error directly adds up at each time step. {\small PMFAST}
did not have this problem since it used CIC interpolation on both meshes.  
This test was performed with CITA256 simulations of very large box size,
which should in principle agree with linear predictions up to the resolution.
The output redshifts are very early, and the upper most curve ($z=10$) was obtained after
about 60 time steps.
We see that deviations are enormous when the random offset is turned off.

\begin{figure}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/power_w_wo_disp_mesh.eps}
  \caption{Dark matter power spectrum, measured at $z=180$, $100$, $20$ and $10$, in a series of CITA256 realizations that are 1000 Mpc/$h$ per side. The dashed line represent the initial condition power spectrum, the dotted lines are the linear predictions, and  the open circles the standard P$^3$M configuration. 
  The dots were obtained by simply removing the random particle offset that is usually applied at each time step. \label{fig:disp_mesh}}
\end{center}
\end{figure}

{\bf (Show how averaging over $~sim10$ time steps reduces the scatter.)}

@@

It is possible to reduce this effect, and thereby improve significantly 
the accuracy of the code, by modifying the value of $r_{max}$, at the cost of increasing the total number of time steps.
Fig. \ref{fig:ra_max} shows a comparison of late time power spectra of a series of CITA256 realizations that originate from the same initial conditions, 
and used the same random seeds to control the fine mesh shifts (mentioned above): only the value of $r_{max}$ was modified between each run. 
We observe that the impact is mostly located in the non-linear regime, where decreasing the time step to 0.006 
allows the simulation to recover about 30 per cent of dimensionless power at the turn over scale, in this simulation configuration.
This gain is greatly affected by the choice of initial redshift, the resolution, and the box size, and ideally one would make
test runs in order to optimize a given configuration.  
As expected, the {\small CPU} resources required to run these simulations increase rapidly as $r_{max}$ decreases, as seen in Table \ref{table:ra_max}. 
In this test case, reducing further at 0.001 shows only a mild improvement in accuracy, but the increase in time is more than a factor of four.
We should mention that with a proper use of the non-Gaussian initial conditions, it is possible to start the simulations at much later redshifts, where the time step sizes are dominated by a stronger mesh force.

### 2012-06-06 paper.tex
Modifications after meeting with Ilian and Garrelt in Stockholm. Lots of edits in the text, new figures, included halo mass function for extended pp_range (nice agreement with ST), added Hugh Merz comments about BlueGene.

\begin{abstract}
This paper presents {\small CUBEP3M}, an upgraded version of {\small PMFAST} that
solves gravity on a two-level mesh and that is among the fastest publicly available N-body codes. 
Among the principal changes, the Poisson solver includes particle-particles interactions
at the sub-grid level and across neighbouring cells, plus each level of the volume decomposition is now cubical,
which allows for massive parallelization of the code.
Force kernel have been improved for enhanced matching of the two mesh contributions near the cutoff length.
We discuss the structure of the code, its accuracy, and its scaling performance.
In addition, many utilities have been added, including a runtime halofinder,
particle identification tags and non-Gaussian initial conditions generator, and we briefly describe their implementation strategy
and accuracy, when applicable .

\end{abstract}

>>>>

\begin{abstract}
This paper presents {\small CUBEP3M}, an upgraded version of {\small PMFAST} that
is now one of the fastest highest performance public N-body code. 
Gravity is solved  on a two-level mesh with sub-grid precision due to the particle-particle (pp)
interactions inside and across neighboring cells.
Among important changes since {\small PMFAST},  {\small CUBEP3M} has a volumetric decomposition,
allowing for a massive parallelization over more than 20,000 cores, and has achieved close to ideal weak scaling
even at this scale.  In addition, force kernels have been improved for enhanced matching of the two mesh contributions near the cutoff length, and the code is now equipped with a unique particle identification tag system that allows reconstruction of individual particle trajectories between time steps.
In parallel, many utilities and extensions have been added to the standard code package, 
including a runtime halo finder, a tunable range for the exact pp force, a non-Gaussian initial conditions generator, etc.
We discuss the structure, the accuracy, any known systematic effects, and the scaling performance
of the code and its utilities, when applicable.
\end{abstract}
